Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-12 manuscript.tex
After some corrections and comments by @valcris

Satellite missions dedicated to measuring the Earths gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.
These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.
Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.
A common approach to achieve this is
to discretize the Earth into tesseroids (Figure~\ref{fig:tesseroid})
instead of rectangular prisms.
However, the integral formula for the gravitational effects of a tesseroid
must be solved numerically.
Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.
Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.
This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.
The Gauss-Legendre Quadrature (GLQ) integration consists of approximating
the volume integral by a weighted sum of the effect of point masses.
An advantage of the GLQ approach is that the accuracy of integration can be
controlled by the number of point masses used.
A disadvantage is the increased computation time as the number of
point masses increases.
Thus, there is a trade-off between accuracy and computation time, a common
theme in numerical methods.
\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.
The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.
However, the question of how to determine the optimal parameters for GLQ
integration remained open.

Previous work by \citet{Ku1977} investigated the use of the Gauss-Legendre
Quadrature in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.
\citet{Asgharzadeh2007} proposed using this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, there has been no investigation if the empirical criteria of
\citet{Ku1977} is valid for gravity gradient components or for tesseroids.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.
The criterion used decide if a division is necessary is
that the minimum distance to the computation point must be larger than
the largest dimension of the tesseroid.
This division is repeated recursively until all tesseroids obey the criteria.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.
These formula are faster to compute and don't have singularities at the poles
like their spherical counterparts.
The Cartesian formula are numerically integrated using a Taylor series
expansion as per \citet{Heck2007}.
\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.
In the so called ``near-zone'' of the computation point they use a finner
discretization with smaller tesseroids.
This is accomplished by dividing the tesseroids along their horizontal
dimensions.
However, the determination of an optimal size of the near-zone remains an
open question \citep{Grombein2013}.

We have implemented a modified version of the adaptive discretion of
\citet{Li2011} into the open-source software package \emph{Tesseroids}.
The software uses the Cartesian formula of \citet{Grombein2013} for improved
performance and robustness.
Previous versions of the software have been used by, e.g.,
\citet{Alvarez2012, Bouman2013, Bouman2013a, Mariani2013, Braitenberg2014,
Braitenberg2011, Fullea2014}.
This article describes the software design and the implementation
of our modified adaptive discretization algorithm.
We also present a numerical investigation of the error committed
in the computations.
These results allow us calibrate the adaptive discretization algorithm
separately for the gravitational potential, gravitational acceleration,
as well as the gravity gradient tensor components.

@@

\begin{equation}
    \iiint\limits_{\Omega}
    f(r^\prime, \lambda^\prime, \phi^\prime)
    d\Omega
    \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    f(r_i, \phi_j, \lambda_k),
    \label{eq:glq3d}
\end{equation}

@@

The accuracy of the integration
depends on the number of point masses used in the summation.
\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.
Figure~\ref{fig:glqerrorsample}
illustrates this effect on the $g_{xy}$ gravity gradient component.
The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.
The maps were calculated on a regular grid
with $100\times100$ points.
Figure~\ref{fig:glqerrorsample}a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two ($2 \times 2 \times 2 = 8$ point masses).
Figure~\ref{fig:glqerrorsample}b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.
Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.
\citet{Ku1977} determined an empirical criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.
Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.
One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.
Figure~\ref{fig:glqerrorsample}c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30
($30 \times 30 \times 30 = 27,000$ point masses).
The computed $g_{xy}$ component more closely resembles
the expect results for a single tesseroid \citep{Asgharzadeh2007}.

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point mases instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30,
        the result is similar to that expected for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

\citet{Li2011} proposed an alternative method
for decreasing the distance between point masses
and achieve an accurate integration.
Instead of increasing the GLQ order,
they keep it fixed to a given number
and divide the tesseroid into smaller parts.
The sum of the effects of the smaller tesseroids
is equal to the gravitational effect of the larger tesseroid.
This division effectively decreases
the distance between nodes because
of the smaller size of the tesseroids.
The criterion for dividing a tesseroid or not
is that the distance to the computation point
must be smaller than a constant times the size of the tesseroid.
This is analogous to the criterion of \citet{Ku1977}
because the size of the tesseroid serves as a proxy
for the distance between point masses.
This procedure is repeated recursively
(i.e., to each of the smaller tesseroids and so on)
until all tesseroids are within the acceptable
ratio of distance and size or a minimum size is achieved.

The advantage of this adaptive discretization is
that the number of point masses is only increased
in regions of the tesseroid that are
closer to the computation point.
Alternatively,
simply increasing the order of the GLQ
would increase the number of point masses
evenly throughout the whole tesseroid.

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates),
as well as model and computation grid generation.
All programs receive input through
command-line arguments and the standard input channel (STDIN)
and output the results through the standard output channel (STDOUT).
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
calculate on it $g_z$ and $g_{zz}$ caused by
the tesseroids in a file ``MODELFILE'',
and save the results to a file called ``OUTPUT''
is:

@@

Our implementation of the adaptive discretization algorithm
differs in a few ways from the original.
In \citet{Li2011},
a tesseroid will be divided when
the smallest distance between it and the computation point
is smaller than a constant times
the largest dimension of the tesseroid.
Instead of the smallest distance,
we use the easier to calculate
distance between
the computation point $(r, \lambda, \phi)$
and the geometric center of the tesseroid
$(r_t, \lambda_t, \phi_t)$

@@

Our definition of the dimensions of the tesseroid
(the ``side lengths'' of \citet{Li2011})
along longitude, latitude, and radius, respectively, are
(Figure~\ref{fig:division}a)

@@

\begin{equation}
    \frac{d}{L_i} < D,
    \label{eq:condition}
\end{equation}

\noindent
for each $i \in (\lambda, \phi, r)$.
$D$ is a positive scalar
hereafter referred to as the ``distance-size ratio''.
If the inequality holds for all three dimensions,
the tesseroid is not divided.
Thus, the distance-size ratio determines
how close the computation point can be
before we must divide the tesseroid.
The value of $D$ is indirectly responsible for
the accuracy of the solution and the computation time.
We will explore the relationship with the accuracy in the following section.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \emph{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ between
        the geometric center of the tesseroid and
        the computation point using equation~\ref{eq:distance}.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number of new elements plus \emph{stktop}
             is smaller than the maximum stack size.
             If it is not (i.e., the stack is full),
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\emph{stktop} is equal to -1).
\end{enumerate}

This stack-based implementation
has some advantages over the original recursive implementation,
namely:
(1) It will be faster because it bypasses the overhead of function calls.
(2) It gives the developer more control over the recursion step.
In recursive implementations,
the developer has no control over
the maximum number of consecutive recursive calls
(i.e., the ``recursion depth'').
This limit may vary with programming language,
compiler, and operating system.
Overflowing the maximum recursion depth
may result in program crashes,
typically with cryptic or inexistent error messages.
In the stack-based implementation,
the developer has complete control.
Overflowing of the stack can be handled gracefully
with an error message
or even performing a suitable approximation of the result.

\subsection{Reproducing the analysis and results}

All figures and numerical experiments in this article
were produced in IPython notebooks\footnote{
\url{http://ipython.org}}
\citep{Perez2007}.
These files combine the analysis and figure generation source code
with text, equations, and the figures generated by the code.
We used the libraries
\emph{pandas} by \citet{Mckinney2010}
to perform the data analysis,
\emph{matplotlib} by \citet{Hunter2007} for graphs,
and \emph{Mayavi} by \citet{Ramachandran2011} for 3D graphs.

The notebook files,
as well as instructions for installing and running the analysis,
are distributed along with
the source code archive that accompanies this article.
Alternatively,
all accompanying material is also available online\footnote{
\url{https://github.com/pinga-lab/paper-tesseroids}}\footnote{
\url{}\todo[inline]{Replace with the Zenodo/figshare DOI for the repository.}}.

@@

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points.
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid and consider only the largest error.
By placing the grid on top of a particular tesseroid, we increase the chances
of capturing the true largest integration error.

>>>>

Satellite missions dedicated to measuring the Earths gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.
These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.
Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.
A common approach to achieve this is
to discretize the Earth into tesseroids (Figure~\ref{fig:tesseroid})
instead of rectangular prisms.
However, the integral formula for the gravitational effects of a tesseroid
must be solved numerically.
Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.
Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.
This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.
The Gauss-Legendre Quadrature (GLQ) integration consists of approximating
the volume integral by a weighted sum of the effect of point masses.
An advantage of the GLQ approach is that it can be
controlled by the number of point masses used.
The larger the number of point masses, the better the accuracy of GLQ
integration.
A disadvantage is the increased computation time as the number of
point masses increases.
Thus, there is a trade-off between accuracy and computation time.
This is a common theme in numerical methods.
\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.
The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.
However, the question of how to determine the optimal parameters for GLQ
integration remained open.

Previous work by \citet{Ku1977} investigated the use of the GLQ in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, the analysis if this ad hoc criteria of
\citet{Ku1977} works well for gravity gradient components or for tesseroids has never been done before.

There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.
A tesseroid is divided if the minimum distance to the computation point
is smaller than the largest dimension of the tesseroid.
\todo[inline]{VAL:
Leo, I have modified the above sentence because I have not understood the original meaning.
}
\todo[inline]{LEO: There was a mistake in that sentence.
I've rewritten it. Is it better now?}
This division is repeated recursively until all tesseroids obey the criterion.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.
These formulas are faster to compute and do not have singularities at the poles
like their spherical counterparts.
The Cartesian formula are numerically integrated using a Taylor series
expansion as per \citet{Heck2007}.
\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.
In the so called ``near-zone'' of the computation point they use a finer
discretization composed by smaller tesseroids.
This is accomplished by dividing the tesseroids along their horizontal
dimensions.
However, the determination of an optimal size of the near-zone remains an
open question \citep{Grombein2013}.

We have implemented a modified version of the adaptive discretion of
\citet{Li2011} into the open-source software package \emph{Tesseroids}.
The software uses the Cartesian formula of \citet{Grombein2013} for improved
performance and robustness.
Previous versions of the software have been used by, e.g.,
\citet{Alvarez2012, Bouman2013, Bouman2013a, Mariani2013, Braitenberg2014,
Braitenberg2011, Fullea2014}.

\todo[inline]{Valeria: Leo,  what does the above sentence mean? Does it mean that these mentioned authors have already used a former version of this software?}
\todo[inline]{Leo: exactly. These are the papers I found that mentioned using Tesseroids.}

This article describes the software design and the implementation
of our modified adaptive discretization algorithm.
We also present a numerical investigation of the error committed
in the computations.
These results allow us calibrate the adaptive discretization algorithm
separately for the gravitational potential, gravitational acceleration,
as well as the gravity gradient tensor components.

@@

\begin{equation}
    \iiint\limits_{\Omega}
    f(r^\prime, \lambda^\prime, \phi^\prime)
    d\Omega
    \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda
    f(r_i, \phi_j, \lambda_k),
    \label{eq:glq3d}
\end{equation}

@@

The accuracy of the integration
depends on the number of point masses used in the summation.
\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.
Figure~\ref{fig:glqerrorsample}
illustrates this effect on the $g_{xy}$ gravity gradient component.
The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.
The maps were calculated on a regular grid
with $100\times100$ points.
Figure~\ref{fig:glqerrorsample}a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two ($2 \times 2 \times 2 = 8$ point masses).
Figure~\ref{fig:glqerrorsample}b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.
Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.
\citet{Ku1977} determined an ad hoc criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.
Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.
One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.
Figure~\ref{fig:glqerrorsample}c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30
($30 \times 30 \times 30 = 27,000$ point masses).
The computed $g_{xy}$ component more closely resembles
the expect results for a single tesseroid \citep{Asgharzadeh2007}.

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point masses instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30. In (c) the horizontal locations of the point masses were not shown. Notice that the result shown in (c) is similar to that expected for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

\citet{Li2011} proposed an alternative method
for decreasing the distance between point masses on the quadrature nodes
aiming at achieving an accurate integration.
Instead of increasing the GLQ order,
they keep it fixed to a given number
and divide the tesseroid into smaller volumes.
The sum of the effects of the smaller tesseroids
is equal to the gravitational effect of the larger tesseroid.
This division effectively decreases
the distance between nodes because
of the smaller size of the tesseroids.
The criterion for dividing a tesseroid is that the distance to the computation point
should be smaller than a constant times the size of the tesseroid.
This is analogous to the criterion proposed by \citet{Ku1977}
because the size of the tesseroid serves as a proxy
for the distance between point masses.
This procedure is repeated recursively
until all tesseroids are within the acceptable
ratio of distance and size or a minimum size is achieved.

The advantage of this adaptive discretization is
that the number of point masses is only increased
in regions where the tesseroid are
closer to the computation point.
Notice that the alternative approach of
simply increasing the order of the GLQ
would increase the number of point masses
evenly throughout the whole tesseroids.

@@

\emph{Tesseroids} consists of command-line programs
written in the C programming language.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates),
as well as model and computation grid generation.
\todo[inline]{Valeria: Leo, I did not understand "as well as model and computation grid generation"}
\todo[inline]{Leo: I mean that there are programs to generate the model files and the grid files. This can be removed without any problem.}
All programs receive input through
command-line arguments and the standard input channel (STDIN)
and output the results through the standard output channel (STDOUT).
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
calculate on it $g_z$ and $g_{zz}$ caused by
the tesseroids in a file "MODELFILE'',
and save the results to a file called "OUTPUT''
is:

@@

Our implementation of the adaptive discretization algorithm
differs in a few ways from one proposed by \citet{Li2011}.
In \citet{Li2011},
a tesseroid will be divided when
the smallest distance between it and the computation point
is smaller than a constant times
the largest dimension of the tesseroid.
Instead of the smallest distance,
we use the easier to calculate
distance between
the computation point $(r, \lambda, \phi)$
and the geometric center of the tesseroid
$(r_t, \lambda_t, \phi_t)$

@@

Our definition of the dimensions of the tesseroid
(the "side lengths'' of \citet{Li2011})
along longitude, latitude, and radius, respectively, are
(Figure~\ref{fig:division}a)

@@

\begin{equation}
    \frac{d}{L_i} > D,
    \label{eq:condition}
\end{equation}
\todo[inline]{Leo: the condition was wrong. It had a $<$ when it should be a $>$. I fixed it.}

\noindent
for each $i \in (\lambda, \phi, r)$.
$D$ is a positive scalar
hereafter referred to as the "distance-size ratio''.
If the inequality holds for all three dimensions,
the tesseroid is not divided.
Thus, the distance-size ratio determines
how close the computation point can be
before we must divide the tesseroid.
The value of $D$ is indirectly responsible for
the accuracy of the solution and the computation time.
We will explore the relationship with the accuracy in the following section.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \emph{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ (equation~\ref{eq:distance}) between
        the geometric center of the tesseroid and
        the computation point.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.

\todo[inline]{  Regarding the Item 5  - I think that it is confusing.

In the item 5 is written " If none of the dimensions fail the above condition, ...."

This  means that all the dimensions hold the inequality (21). Ok ?

So, Item 5 could be written directly,

5. If all the dimensions hold the inequality (21), hence the tesseroid is not divided and the gravitational effect of the tesseroid is computed by using the
        Gauss-Legendre Quadrature.....}
\todo[inline]{Leo: I agree completely. I have made the modifications below.}

    \item If all dimensions hold the inequality \ref{eq:condition},
        the tesseroid is not divided and its
        gravitational effect is computed using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.

\todo[inline]{   Regarding the Item 6b  - I am not sure, but I think that it might be wrong or written in an awkward style

Item 6b is written - "Check if the number of new elements plus \emph{stktop} is smaller than the maximum stack size."
Based on item 6c, I think that in 6b you are check if the stack is full, ok ?

My suggestion is to merge items 6b and 6c, for example:

If the stack is full, place the smaller tesseroids into the stack; otherwise warn the user of "stak overflow" and compute the effect of the larger tesseroid (step 5)

}
\todo[inline]{Leo: Yes, that sentence was a bit confusing.
The idea is checking if the newly generated tesseroids
fit in the stack.
I just wanted to explain how to check for that
(the number of new elements ...).
I'll rephrase it.}
\todo[inline]{Leo: By the way, the proper way of writing " in LaTeX is ``like this''.}

        \item Check if there is room in the stack
            for the new tesseroids
            (i.e.,the number of new elements plus \emph{stktop}
             is smaller than the maximum stack size).
             If there isn't,
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
             If there is room in the stack,
             place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\emph{stktop} is equal to -1).
\end{enumerate}

This stack-based implementation
has some advantages over the original recursive implementation,
namely:
(1) It will be faster because it bypasses the overhead of function calls.
(2) It gives the developer more control over the recursion step.
In recursive implementations,
the developer has no control over
the maximum number of consecutive recursive calls
(i.e., the "recursion depth").
This limit may vary with programming language,
compiler, and operating system.
Overflowing the maximum recursion depth
may result in program crashes,
typically with cryptic or inexistent error messages.
In the stack-based implementation,
the developer has complete control.
Overflowing of the stack can be handled gracefully
with an error message
or even performing a suitable approximation of the result.

\subsection{Code for figures and error analysis}

The error analysis and all figures in this article
were produced in IPython notebooks
\citep{Perez2007}.
The notebook files combine source code in various programming languages,
program execution,
text, equations,
and the figures generated by the code
into a single document.
We used the following Python language libraries
to perform the error analysis and generate figures:
\emph{pandas} by \citet{Mckinney2010},
\emph{matplotlib} by \citet{Hunter2007} for 2D figures and maps,
and \emph{Mayavi} by \citet{Ramachandran2011} for 3D figures.

\todo[inline]{ VAL: The code was written in C., was not it? I think that there is a  version in Python. However, the readers do not known. }

\todo[inline]{Leo: The code that we are discussing in this paper was written in C.
The figures were generated in IPython notebooks in Python.
The error analysis was done in IPython notebooks as well.
However, it does NOT use the Python version of the tesseroid code.
It runs the C programs and uses Python to calculate the differences
The modeling that is written in Python is not included or discussed here.
Readers do not need to know that it exists.}

\todo[inline]{VAL: The above sentence is confusing:
1) "These files combine the analysis and figure.... ",
What are the files? Probably, they are the IPython notebooks. Ok?
Why are you saying "the analysis"
In the last paragraph, you said "running the analysis".
What is the analysis? Is the test to evaluate the accuracy?
I think that you want to say
"The Ipython nootebook files combine source codes in python, code executions, text, equations, dnd the figures generated by the code into a single document"
You need keeping in mind that you know about Ipython nootebook. However, the readers cannot know.}

\todo[inline]{LEO: Yes, that part is a bit confusing now that I read it again.
By analysis, I mean the error analysis.
That is done in \href{http://nbviewer.ipython.org/github/pinga-lab/paper-tesseroids/blob/master/notebooks/tesseroid_vs_spherical_shell.ipynb}{this IPython notebook}.
This file will be  available  with the article.
There is a link to this online version in \url{https://github.com/pinga-lab/paper-tesseroids}
which is mentioned in the text.
There are two other notebooks that make the figures for the methods and the final results.
There are links to those two in the above link  as well.
I'll rewrite the paragraph to make it more clear.}

The IPython notebooks
and the data generated for the error analysis,
as well as instructions for installing the software
and running the programs,
are also included in
the source code archive that accompanies this article.
Alternatively,
all accompanying material is available
in online repositories\footnote{
\url{https://github.com/pinga-lab/paper-tesseroids}}\footnote{
\url{}\todo[inline]{Replace with the Zenodo/figshare DOI for the repository.}}.

@@

\todo[inline]{ VAL: The above paragraph is great !}

@@

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.
Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.
Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points at different geographic locations
(Table~\ref{tbl:experiment}).
\todo[inline]{ VAL: A grid of 10 X  10  point covers a are of ??? X ???? km}
\todo[inline]{LEO: The grid location varied in each experiment of the numerical analysis. Table 1 shows the locations of the grids.
This is discussed below.
I'll add this information to the text.}
Fortunately, the symmetry of the shell allows us to consider the computation
point at any geocentric coordinate.
Therefore, the effect of the shell will be same along the entire grid.
We compute the differences between the effects of the shell and the tesseroid
model on the grid.
However, we will consider only the largest error in our analysis.
\todo[inline]{ VAL: I rewrite the above sentence by:
"We compute the differences between the effects of the shell and the tesseroid model on the grid; however, we consider  the largest error only.  }
\todo[inline]{LEO: Rewritten. Is it better now?}
We placed the grid on top of a particular tesseroid
to increase the chances of capturing the true largest integration error.
\todo[inline]{ VAL:I rewrite the above sentence as:
"We decided for the nonconservative procedure of placing the grid on top of a particular tesseroid to increase the chances
of capturing the true largest integration error"}
\todo[inline]{LEO: Changed. I don't think this is really
``nonconservative'' so I left it out.}

\todo[inline]{ I've stopped here - VALERIA}
